---
  title: "Analysis Of Policing Dataset From Dallas, Texas in 2016."
author: "Shekar, Mahesh"
output: 
  html_document: default
---
  
  ## INTRODUCTION
  The purpose here is to discover trends or patterns in police activity, this report will present an in-depth research of the policing practises in Dallas, Texas in 2016. By analysing the data, we intend to identify any possible problems and areas for development and to make suggestions for policy changes that might result in more equal and efficient police methods.<br>
  The objective here is to examine the police activity, arrests, use of force and other activities from Dallas, Texas dataset in 2016 and look for any patterns or trends. The report will also seek to pinpoint any racial, gender based discrepancies in policing practises and offer suggestions for change.

## Steps involved in analysis of data visualization:
* Data cleaning and preparation
* Exploratory data analysis
* Make appropriate visualizations
* Interpret your findings
* Communicate results

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
#Set work directory
setwd("C:/Users/HP/OneDrive/Desktop/MA304-Data Visualisation/Completed")


#Read the csv file
proj_dallas <- read.csv("37-00049_UOF-P_2016_prepped.csv")


#Install requried libraries
library(dplyr)
library(leaflet)
library(viridis)
library(parsedate)
library(tidyr)
library(stringr)
library(ggplot2)
library(kableExtra)
library(tidygeocoder)
library(plotly)


#Clean the data by: 
#Correcting the date format 
proj_dallas <- proj_dallas %>% mutate(INCIDENT_DATE = parse_date(INCIDENT_DATE),
                                      OFFICER_HIRE_DATE = parse_date(OFFICER_HIRE_DATE))


#Separating rows having multiple values
proj_dallas <- proj_dallas %>% separate_rows(UOF_NUMBER, FORCE_EFFECTIVE, sep = ",")


#Trimming the space from left and right of the values
proj_dallas <- proj_dallas %>% mutate(across(where(is.character),str_trim))


#Convert char to factor
proj_dallas <- proj_dallas %>%
  mutate_at(c('OFFICER_GENDER','OFFICER_RACE','OFFICER_INJURY','OFFICER_INJURY_TYPE',
              'OFFICER_HOSPITALIZATION','SUBJECT_RACE','SUBJECT_GENDER','SUBJECT_INJURY',
              'SUBJECT_INJURY_TYPE','SUBJECT_WAS_ARRESTED','SUBJECT_DESCRIPTION','SUBJECT_OFFENSE',
              'DIVISION','INCIDENT_REASON','REASON_FOR_FORCE','FORCE_EFFECTIVE'),as.factor)


#Reducing the columns for 'type_of_forced' used by rows
proj_dallas <- proj_dallas %>% pivot_longer(c("TYPE_OF_FORCE_USED1","TYPE_OF_FORCE_USED2","TYPE_OF_FORCE_USED3",
                                              "TYPE_OF_FORCE_USED4","TYPE_OF_FORCE_USED5","TYPE_OF_FORCE_USED6",
                                              "TYPE_OF_FORCE_USED7","TYPE_OF_FORCE_USED8","TYPE_OF_FORCE_USED9",
                                              "TYPE_OF_FORCE_USED10"), names_to = "TYPE_OF_FORCE",
                                            values_to = "ACTIONS_TAKEN")


#Filter Out empty rows
proj_dallas <- proj_dallas %>%
  filter(ACTIONS_TAKEN != "")


#Remove unwanted column
proj_dallas <- proj_dallas[,-c(25:27,30,31,36), drop = FALSE]
```

## EXPLORING VISUALIZATION
After cleaning and filtering out the data we are now exploring the data and finding out patterns and observations through visualization.<br>
  
  ## RESULTS
  
  ### 1.Displaying **Arrest** of subjects in Dallas
  ```{r echo=FALSE, message=FALSE, warning=FALSE, comment=NULL}
location_data <- proj_dallas %>%
  filter(!is.na(LOCATION_LATITUDE), SUBJECT_WAS_ARRESTED == "Yes") %>%
  group_by(SUBJECT_ID, LOCATION_LATITUDE, LOCATION_LONGITUDE, LOCATION_DISTRICT) %>%
  summarise(count = n()) %>%
  filter(SUBJECT_ID != 0, LOCATION_DISTRICT != "NULL")

my_loc_palette <- colorFactor(palette = viridis(7),
                              domain = location_data$LOCATION_DISTRICT,
                              na.color = "transparent")

leaflet(location_data) %>%
  setView(lng = -96.808891, lat = 32.779167, zoom = 10) %>%
  addProviderTiles(providers$Stamen.TonerLite) %>%
  addCircleMarkers(
    lng = ~LOCATION_LONGITUDE,
    lat = ~LOCATION_LATITUDE,
    fillColor = ~my_loc_palette(location_data$LOCATION_DISTRICT),
    fillOpacity = 0.8,
    color="white",
    radius=8,
    stroke=F,
    label=~paste("Arrested: ", count),
    popup=~paste("Arrested: ", count, "\nLOCATION_DISTRICT: ", LOCATION_DISTRICT)
  ) %>%
  addLegend(pal=my_loc_palette, values = ~LOCATION_DISTRICT, title = "Regions", 
            position = "bottomleft",
            opacity = 1)
```
<br>
  The legend on the bottom-left of the map shows the color-coding of the police divisions. "Regions" is the name of the legend, and it lists the colors that are utilized for each district. Users can zoom in and out of the interactive map and hover their cursors over the circle markers to view popups.<br>
  Based on the police division where the arrest was made, the map is color-coded. The size of each circle on the map denotes an arrest site and corresponds to the number of arrests that occurred there. Additionally, there are labels on the circle markers that list the number of arrests that have taken place there, and a popup window that displays both that information.<br>
  The graph shows where police are making arrests in Dallas, Texas, in terms of location. Additionally, it reveals which police departments are making the most arrests, which is helpful for analyzing policing trends and identifying problematic zones.
<br>
  
  ### 2.Count of highest **Arrest** by **Divison**
  ```{r echo=FALSE, message=FALSE, warning=FALSE, comment=NULL}
#Highest crimes in location
proj_dallas %>% select(SUBJECT_WAS_ARRESTED, DIVISION) %>%
  filter(SUBJECT_WAS_ARRESTED == "Yes") %>%
  group_by(DIVISION) %>% count() %>%
  ggplot(aes(x = DIVISION, y = n)) +
  geom_histogram(stat = "identity", fill = "violet")
```
<br>
  There are 7 subsets under DIVISION in that according to the above graph CENTRAL has the highest count of arrest compared to the others, which have almost the same count of arrest. The CENTRAL looks like the center of the Town so eventually have more people activities. And it is observed that in general these are the places where crimes happen.
<br>
  
  ### 3.which month has **Highest Arrest** in 2016.
  <br>
  ```{r echo=FALSE, message=FALSE, warning=FALSE, comment=NULL}
#which month has highest robbery
proj_dallas %>% select(INCIDENT_DATE, SUBJECT_WAS_ARRESTED) %>%
  filter(SUBJECT_WAS_ARRESTED == "Yes") %>%
  mutate(month = format(INCIDENT_DATE, "%m")) %>%
  group_by(month) %>%
  summarise(count_arrest = n()) %>%
  ggplot(aes(x = month, y=count_arrest)) +
  geom_col(fill = "orange")
```
<br>
  Here, the generated graph displays the monthly total of arrests for each incidence in the data set. The graph demonstrates that, except for February and March, the number of arrests is largely stable from month to month. This may have been to various reasons or just a coincident, highlighting instances where the subject was taken into custody. Even though the analysis's reach is limited, it offers some insights on the trends in Dallas's arrests and may serve as the basis for future research.
<br>
  
  ### 4.Count of officers **Injured** 
  ```{r echo=FALSE, message=FALSE, warning=FALSE, comment=NULL}
#count of officers injuried
officers_injuried <- proj_dallas %>% select(OFFICER_GENDER, OFFICER_HOSPITALIZATION) %>%
  group_by(OFFICER_GENDER, OFFICER_HOSPITALIZATION) %>% count() %>%
  pivot_wider(names_from = OFFICER_HOSPITALIZATION, values_from = n)
officers_injuried
```
<br>
  The number of police officers injured by gender and hospitalized status is displayed in a table. 'No' and 'Yes' are the two columns in the table that show whether the patient was hospitalized. 'Female' and 'Male' rows denote the officers' genders. The table shows that 386 Male and 45 Female officers were hurt during the arrest of the subject and were hospitalized. By this we can assume that the subjects were armed or were not peacefully willing to surrender themselves.
<br>

#### Minor and Major injuries
```{r echo=FALSE, message=FALSE, warning=FALSE, comment=NULL, fig.width=5, fig.height=5}
#minor and major injuries
proj_dallas %>% select(OFFICER_GENDER, OFFICER_INJURY_TYPE,OFFICER_HOSPITALIZATION) %>%
  filter(OFFICER_INJURY_TYPE != "No injuries noted or visible" &
           OFFICER_HOSPITALIZATION == "Yes") %>%
  group_by(OFFICER_GENDER) %>% count() %>%
  ggplot(aes(x = "", y = n, fill = OFFICER_GENDER)) +
  geom_bar(stat = "identity", width = 1) +
  geom_text(aes(label = paste(round((n/sum(n))*100, 2), "%")),
            position = position_stack(vjust = 0.5)) +
  coord_polar("y", start = 0)+
  theme_void()
```
<br>
The percentage of male and female officers who required hospitalization due to injuries is displayed in the pie chart. Male and female officers are represented by the left and right halves, respectively, of the chart. According to the chart, male officers were hospitalized for injuries at a higher rate than female officers. This results on the hard, dangerous and critical situations a officer has to face to apprehend a subject.
<br>

### 5. People **Joining Force** over the years
```{r echo=FALSE, message=FALSE, warning=FALSE, comment=NULL, fig.width=12}
#how many and which year who has joined more
proj_dallas %>%
  mutate(year = format(OFFICER_HIRE_DATE, "%Y")) %>%
  group_by(year) %>%
  count() %>%
  ggplot(aes(x = year, y = n, group = "red")) +
  geom_line(aes(color = "red")) +
  geom_point(aes(color = "red")) +
  theme_bw() +
  theme(legend.position = "none")

```
<br>
Chart shows a trend in the number of officers hired over the years. The chart shows that the number of officers hired has been increasing gradually from 2005 to 2016. As time has passed people have shown interest in joining the force, this shows the police force has been increased since the rise in the population vice versa crimes and but not to forget the interest, determination and confidence of the people in the force.    
<br>

### 6.People of different **Race** joining force 
```{r echo=FALSE, message=FALSE, warning=FALSE, comment=NULL, fig.width=12, fig.height=12}
proj_dallas %>% select(OFFICER_ID, OFFICER_RACE, OFFICER_HIRE_DATE) %>%
  mutate(year = format(OFFICER_HIRE_DATE, "%Y")) %>%
  group_by(year, OFFICER_RACE) %>%
  count() %>%
  ggplot(aes(x = year, y = n)) +
  geom_col(fill = "turquoise4") +
  coord_flip() +
  facet_wrap(vars(OFFICER_RACE), scales = "free") +
  theme_classic()
```
<br>
By year and race, the number of cops hired is represented on the graph. The y-axis shows the number of police hired, the x-axis the years, and the bars show the number of officers hired annually for each race.<br>
From the graph we can see that there is a high number of counts from both black and white race people, compared to all the other races. But the white race people have the highest count for joining the force since 1979. It is observed that American Ind has shown a sudden spike in 2012 and later have not joined the force at all. This comes for the Asian people too, where there is a sudden spike in 2014 compared to other year and 2015 is the last year, they have joined the force and among the other categories too, this may be because they lacked the interest to join the force or there may have been new rules and regulations passed by the force where they were not eligible to join the force. We cannot come to that exact conclusion since we lack the data.
<br>

### 7.Officers **Time** spent in force
```{r echo=FALSE, message=FALSE, warning=FALSE, comment=NULL}
#officers years spent on force
proj_dallas %>% select(OFFICER_ID,OFFICER_YEARS_ON_FORCE) %>%
  group_by(OFFICER_YEARS_ON_FORCE) %>% count() %>%
  ggplot(aes(x = OFFICER_YEARS_ON_FORCE, y=n)) +
  geom_density(stat = "identity", fill = "green")
```
<br>
The y-axis displays the density of policemen for each year of service, while the x-axis displays the number of years the police officers have been on the force. The graph's density distribution of officers by number of years of service is shown in green.<br>
  The graph shows that most of the Dallas police force has been in service for less than ten years. There are more officers with fewer years of service than those with more, according to the density distribution, which looks to be skewed to the right.<br>
  Overall, this graph gives a great visualization of the years of service distribution among Dallas's police officers, which may be beneficial for decisions regarding police officer hiring, training, and retention.
<br>

### 8.Officers time spent in force based on **Race**
```{r echo=FALSE, message=FALSE, warning=FALSE, comment=NULL}
#boxplot
proj_dallas %>% group_by(OFFICER_RACE, OFFICER_YEARS_ON_FORCE) %>%
  ggplot(aes(x = OFFICER_RACE, y = OFFICER_YEARS_ON_FORCE, fill = OFFICER_RACE)) +
  geom_boxplot(show.legend = FALSE)
```
<br>
The y-axis displays the amount of years the officers have been on the force, while the x-axis shows the various officer races. The distribution of years of service for officers of a specific race is shown in each boxplot.<br>
From the graph it can be observed that, the median years of service are almost similar across all officer race expect for Asian officers who have higher median years of service compared to others. However for white and Hispanic there is a wider range of years of service compared to other officers race.<br> 
A few outliers in the data also seem to exist, especially among black officers who have been in the service for a longer period than officers of other races.
<br>

### 9.Corelation b/w **Type of Forced Used** by Officers based on **Years of Experience**
```{r echo=FALSE, message=FALSE, warning=FALSE, comment=NULL}
#correlation
df1 <- proj_dallas %>%
  group_by(OFFICER_ID, TYPE_OF_FORCE) %>%
  summarise(count = n()) %>%
  select(c("OFFICER_ID", "count")) %>%
  filter(OFFICER_ID != 0)

df1 <- df1[!duplicated(df1), ]

df1 <- df1 %>%
  group_by(OFFICER_ID) %>%
  summarise(total = sum(count)) %>%
  arrange(OFFICER_ID) %>%
  select(total)

corr_force_officer <- cbind(df1,
                            proj_dallas %>%
                              select(c(OFFICER_ID, OFFICER_YEARS_ON_FORCE)) %>%
                              group_by(OFFICER_ID) %>%
                              slice(1) %>%
                              ungroup %>%
                              filter(OFFICER_ID != 0)
                            )

cor_graph <- corr_force_officer %>% ggplot(aes(x = total, y = OFFICER_YEARS_ON_FORCE)) +
  geom_point(color = "pink") +
  geom_smooth()

ggplotly(cor_graph)
```
<br>
The scatterplot with a trend line in the graph above illustrates the relationship between the level of physical actions taken by Dallas police officers and their length of service. An officer's overall use of force is represented by the x-axis, and the number of years they have been on the force is represented by the y-axis.<br>
  Officers with more years of service are unlikely to use force more frequently than those with less years of service, according to the trend line, which indicates a very weak correlation between the two factors. The fact that several data points deviate from the trend line, showing a substantial level of variation in the data, should be noted.<br>
  Overall, this graph provides some light on the correlation between Dallas police officers' use of force incidents and years of service. In order to fully understand the elements that lead to the use of force by police officers and how this may differ by years of service, further study would be required. It's essential to consider the data limitations as well as the possibility that other factors, such as race and gender, may affect how much force is used by police.<br>
  
  ## CONCLUSION
  There are chances for improvement in the way police officers carry out their jobs, according to the data analysis and visualization of policing practices in Dallas, Texas, in 2016.<br>
  We’ve seen that the Central police division had the most arrests, which shows that this area needs to be watched over more carefully. More funding should be given to this division to enable better oversight of police operations.<br>
  Second, the data showed that male police officers suffer injuries more frequently than their female counterparts, suggesting that police agencies should increase training on how to deal with risky situations. Further investigation is also required to identify the causes of these injuries and discover solutions.<br>
  Thirdly, the graph showing the number of arrests by month indicates that, apart from February and March, the number of arrests is largely steady throughout the year, suggesting that there may be some external variables that affect the number of arrests made.<br>
  Fourth, the data demonstrates a favorable trend in which the number of officers hired increased steadily between 2005 and 2016. There may be certain institutional hurdles to their recruitment, as evidenced by the fact that the number of officers hired from specific races—such as American Indian and Asian—is much lower than that of other races. Investigating this further is necessary.<br>
  Finally, the density distribution graph showing the years of service of police officers reveals that most of them have been on the force for less than 10 years, which may indicate a lack of experience and expertise in some areas. It is advised that experienced police train and guide new recruits more.
In conclusion, various opportunities for improvement were identified throughout the analysis of police practices in Dallas, Texas in 2016. We hope that by highlighting these issues, decision-makers may develop more equitable and effective police strategies.<br>
  